Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-17 _tex/paper2.tex
Built site for gh-pages

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} helps researchers to
determine whether an estimator can accurately compute an estimand based
solely on its (causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different teaching methods influence students' writing
ability. Each structural component defines how observed or latent
variables affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses considering the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}) and the uncertainties in
trait estimation (see Section~\ref{sec-theory-issue2}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
latent vectors of length \(n_{I}\). Additionally, this branch shows how
\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
text-related variables \((X_{IA})\) (e.g., text length), and texts'
idiosyncratic errors \(e_{IA}\) causally influence the texts'
written-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
judges' biases associated with each text \(B_{JK}\), the second primary
latent variable of interest. The error \(e_{JK}\) captures variations in
judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
matrices with \(n_{J}\) rows and \(n_{K}\) columns

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. In
Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
researchers work with a data sample consisting of a limited number of
repeated judgments \((n^{s}_{K})\) from a sample of judges
\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
sample of students \((n^{s}_{I})\), all drawn from the conceptual
population. Furthermore, in Section~\ref{sec-theory-theoretical_SC2},
the model assumes that judges do not perform \emph{all repeated
judgments} within the data sample. Instead, they conduct a sufficient
number of stimuli comparisons, \(n_{C}\), to ensure an accurate
estimation of the proportion \(P(B>A)\), as proposed by
\citet{Thurstone_1927b}.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model (see equation
(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
the number of rows in the \(R\) matrix, with each element
\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
absence of data rows in \(R\), a definition similar to that of
\(s_{(i,a,h,b,j,k)}\) in equation (\ref{eq-mat26}).
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
directly access to \(O^{sc}_{R}\) due to the comparison mechanism.
Moreover, the DAG depicts \(C\) as a causal factor influencing
\(O_{R}\), enclosing it in a square to indicate that it is a conditioned
variable. Specifically, \(C\) determines \emph{which repeated judgments
the subset of judges makes for the subset of stimuli produced by the
sampled individuals}. In essence, \(C\) reflects the assumption that
judges \emph{do not} perform all possible repeated judgments for the
selected stimuli. Instead, they carry out a ``sufficient number of
comparisons'' \(n_{C}\) to allow for accurate estimation of the
proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-issue1b}), reflect the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and incorporate
measurement error in trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, they do not resolve concerns
about the assumption of equal dispersions among stimuli discussed in
Section~\ref{sec-theory-issue1a}. Because this issue involves
statistical assumptions about the distribution of the discriminal
process, it requires a statistical model, which we develop in the next
section.

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
appropriate functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
  binomial distribution--including its special case, the Bernoulli
  distribution-- represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that when only two
  un-ordered outcomes are possible, and the expected numbers of each
  type of event are assumed to be constant, then the binomial
  distribution is the one that is most consistent with these constraints
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that while the
probabilistic model \ref{fig-cj16_prob} includes the probability
distributions for the sampling and comparison mechanisms, \(P(S)\) and
\(P(C)\), we omit these from the statistical model \ref{fig-cj16_stat}.
We do so because, although these distributions contribute to the joint
distribution of the data, they do not affect the relationship between
\(O_{R}\) and \(D_{R}\). This lack of influence is due to the random
selection designs that govern \(S\) and \(C\)\footnote{Randomization
  ensures that data---and, by extension, an estimator---satisfies
  several key identification properties, such as common support, no
  interference, and consistency. The most critical property, however, is
  the elimination of confounding. \emph{Confounding} occurs when an
  external variable, such as \(X_{I}\), simultaneously influences both
  the outcome (e.g., \(O_{R}\)) and a variable of interest (e.g.,
  \(S\)), resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

@@

\[
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
\]

@@

@@

Sixth, we define \(B_{J}\) as a linear combination of the effects of
relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
which each judge belongs. For instance, if \(Z_{J}\) is a variable that
can take on three values representing three groups of judges with
varying expertise, the model imposes the constraint
\(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the scale of the latent trait
\citep{Depaoli_2021}.

Finally, to transform model~\ref{fig-cj16_stat} into a practical
statistical tool for analyzing paired comparison data, we use
\emph{Bayesian inference methods}. The \textbf{Declarations} section of
this document includes a link to the model, along with an alternative
specification that assumes equal discriminal dispersions. We implemented
both versions of the model using \texttt{Stan} \citep[version
\(2.26.1\),][]{Stan_2020}. Although the implementation was successful,
we do not report the results here, as they fall outside the scope of
this document.

>>>>

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different teaching methods influence students' writing
ability. Each structural component defines how observed or latent
variables affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. The model's hierarchical design allows
researchers to formulate and test hypotheses that account for both the
nested structure of stimuli and the uncertainties inherent in trait
estimation (see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for further discussion of these
considerations).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
  to footnote 2.} text-related variables \((X_{IA})\) (e.g., text
length), and texts' idiosyncratic errors \(e_{IA}\) causally influence
the texts' written-quality trait \((T_{IA})\), the first primary latent
variable of interest. The error term \(e_{IA}\) captures variations in
the texts' traits that remain unexplained by \(T_{I}\) or \(X_{IA}\).
Here, \(X_{IA}\) is an observed matrix with dimensions
\(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) independent columns
(variables), while \(e_{IA}\) and \(T_{IA}\) are latent matrices with
\(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
independent columns (variables), and both \(e_{J}\) and \(B_{J}\) are
latent vectors of length \(n_{J}\). Furthermore, the branch shows how
\(B_{J}\), along with \emph{relevant} \footnote{refer to footnote 2.}
judgment-related variables \((Z_{JK})\) (e.g., the number of judgments a
judge makes), and judgments' idiosyncratic errors \(e_{JK}\) causally
influence the judges' biases associated with each text \(B_{JK}\), the
second primary latent variable of interest. The error \(e_{JK}\)
captures variations in judgments unexplained by \(B_{J}\) or \(Z_{JK}\).
Here, \(Z_{JK}\) is an observed matrix with dimension
\(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) independent columns
(variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent matrices
with \(n_{J}\) rows and \(n_{K}\) columns

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. First, in
Section~\ref{sec-theory-theoretical_SC1}, it explicitly assumes
researchers work with a data sample consisting of a limited number of
repeated judgments \((n^{s}_{K})\) from a sample of judges
\((n^{s}_{J})\) and a specific number of texts \((n^{s}_{A})\) from a
sample of students \((n^{s}_{I})\), all drawn from the conceptual
population. Second, in Section~\ref{sec-theory-theoretical_SC2}, the
model assumes that judges do not perform \emph{all repeated judgments}
within the data sample. Instead, they conduct a sufficient number of
stimuli comparisons, \(n_{C}\), to ensure an accurate estimation of the
proportion \(P(B>A)\), as proposed by \citet{Thurstone_1927b}.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model. Equation (\ref{eq-mat27})
shows that \(C\) contains \(n\) elements corresponding to the number of
rows in the \(R\) matrix, with each element \(c_{(i,a,h,b,j,k)}\) being
a binary value indicating the presence or absence of data rows in \(R\),
a definition similar to that of \(s_{(i,a,h,b,j,k)}\) in equation
(\ref{eq-mat26}). \begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
as unobserved, highlighting that researchers cannot directly access to
\(O^{sc}_{R}\) due to the comparison mechanism. Moreover, the DAG
depicts \(C\) as a causal factor influencing \(O_{R}\), enclosing it in
a square to indicate that it is a conditioned variable. Specifically,
\(C\) determines \emph{which repeated judgments the subset of judges
makes for the subset of stimuli produced by the sampled individuals}. In
essence, \(C\) reflects the assumption that judges \emph{do not} perform
all possible repeated judgments for the selected stimuli. Instead, they
carry out a ``sufficient number of comparisons'' \(n_{C}\) to allow for
accurate estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

@@

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), reflect the hierarchical
structure of stimuli (see Section~\ref{sec-theory-theoretical_P3}),
incorporate measurement error in trait estimation and hypothesis testing
(also see Section~\ref{sec-theory-theoretical_P3}), and clarify the role
of the sample and comparison mechanisms in CJ experiments (see
Section~\ref{sec-theory-theoretical_SC}). However, they do not resolve
concerns about the assumption of equal dispersions among stimuli
discussed in Section~\ref{sec-theory-issue1a}. Because this concern
relates to the statistical assumption underlying the distribution of the
discriminal process, we develop a formal statistical model to address it
in the next section.

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in the SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_scm}, and the statistical model \ref{fig-cj16_scm}. First,
we assume that \(O_{R}\) follows a Bernoulli distribution\footnote{The
  binomial distribution--including its special case, the Bernoulli
  distribution-- represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that when only two
  un-ordered outcomes are possible, and the expected numbers of each
  type of event are assumed to be constant, then the binomial
  distribution is the one that is most consistent with these constraints
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

\[
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
\]

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data---and, by extension,
  an estimator---satisfies several key identification properties, such
  as common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(X_{I}\) or \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

@@

Sixth, we define \(B_{J}\) as a linear combination of the effects of
relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., the effects of
judgment expertise on the quality assessment) and judge-specific
idiosyncratic errors \(e_{J}\). Each element of \(e_{J}\) is likewise
assumed to be \(iid\) Normal with mean zero and a standard deviation
\(s_{ZJ}\) that depends on the groups to which each judge belongs. For
instance, if \(Z_{J}\) is a variable that can take on three values
representing three groups of judges with varying expertise, the model
imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
scale of the latent trait \citep{Depaoli_2021}.

Finally, using \emph{Bayesian inference methods} to transform the
model~\ref{fig-cj16_stat} into a practical statistical tool for
analyzing paired comparison data. Bayesian inference offers three key
advantages. First, it is well-suited for estimating complex and
overparameterized models, where the number of parameters may exceed the
number of observations \citep{Baker_1998, Kim_et_al_1999}. Second, it
allows researchers to incorporate prior information, which constrains
parameter estimates within plausible bounds. This feature mitigates
common estimation issues, such as non-convergence or improper solutions,
often encountered in complex models under frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. Third, Bayesian inference
is particularly effective for drawing conclusions from small samples,
where the asymptotic properties underlying frequentist methods are less
reliable \citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}.
The \textbf{Declarations} section of this document provides a link to
the model code, along with an alternative specification that assumes
equal discriminal dispersions. We implemented both versions of the model
using \texttt{Stan} \citep[version 2.26.1]{Stan_2020}. Although the
implementations were successful, we do not report the results here, as
they fall outside the scope of this document.
